---
title: "Medicines Used in Mental Health"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

  <style type="text/css">

  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# England April 2017 to `r max_filter_12_months` {.toc-ignore}

Published 8 March 2023

``` {r changes_box}

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have rewritten the statistical summary narrative to be shorter. We have also made changes to the methodology of the model used in predicting items based on prescribing trends from before the Covid-19 pandemic. We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`

```

## Key findings

Of the five BNF sections covered in these statistics, antidepressants remained the section with the largest number of items and patients in primary care prescribing. Between October to December 2022, 21.7 million antidepressants were prescribed to an estimated 6.59 million identified patients. However, these were proportionally small increases compared to the previous three months, with items increasing by 1.34% and patients by 0.71%. CNS stimulants and drugs for ADHD increased by 6.28% over the same period to 646,000 items, the largest percentage change of all sections in this publication. Antipsychotics and drugs for dementia also showed small increases.

Hypnotics and anxiolytics were the only group of drugs that did not increase between October to December 2022. They showed a negligible decrease of 0.12% to 3.44 million items, so overall remained similar to the previous quarter. Patients decreased slightly more by 0.79%, but still remained above one million patients.

All sections saw a small decrease in both items and identified patients from November to December 2022. This is the first time since 2018 that items and identified patients have both decreased during this time of year across all sections.

In the 34-month period since the implementation of lockdown measures during the coronavirus (COVID-19) pandemic, between March 2020 and December 2022:

* 

* 

---

## 1. Introduction
### 1.1. Scope {.toc-ignore} 

The [NHS Long Term Plan](https://www.longtermplan.nhs.uk/) published in 2019 focuses on improving mental health care in the UK. The Long Term Plan aims to provide more funding for mental health services, particularly for children and people with dementia. It also seeks to improve community care for those with severe mental illnesses as part of the [Core20PLUS5](https://www.england.nhs.uk/about/equality/equality-hub/core20plus5/) approach to reduce health inequalities.

This publication provides information on the medicines used to treat mental health conditions in England. The information included only covers medicines prescribed in England that then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands by a pharmacy, appliance contractor, dispensing doctor, or have been personally administered by a GP practice. The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors. Is also discusses how the COVID-19 pandemic has affected the use of these medicines.

The publication uses the British National Formulary (BNF), a list of medicines that is used in the UK to classify medicines according to their primary therapeutic use. Medication may also be prescribed for other purposes beyond their intended use. For instance, certain antidepressants may be recommended to individuals experiencing chronic pain. The NHSBSA does not capture the clinical indication of a prescription during processing. The publication focuses on five sections of the BNF that cover medicines used to treat anxiety, depression, psychosis, attention deficit hyperactivity disorder (ADHD), and dementia.

It’s important to note that the publication does not discuss the costs associated with the provision of there medicines. However, this information is available in the supporting summary tables which accompany this release. Additionally, it’s worth noting that medicines are not the only way to treat mental health conditions, and this publication does not capture information on the use of other metal health services, such as psychological therapies. As a result, this publication may not give a complete picture of the number of people receiving treatment for mental health conditions.

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS Digital mental health data hub] (https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide further context. These include a mental health bulletin, reports on children and young people's mental health surveys, dementia assessment and referral statistics, and more.

### 1.2. Definitions {.toc-ignore}

``` {r definition_infoboxes}

::::{.row}
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("Item", text = "A single unit of mediciation listed separately on a prescription form.", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("Patient", text = "A unique NHS number verified from a prescription form.", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("NIC", text = "The Net Ingredient Cost (NIC) is the basic price of the medication and the quantity prescribed", width = "100%")`
:::
::::

```

In this publication, an example of an item would be Fluoxetine 20mg tables x56.

Verification of an NHS number for patient identification is done by the NHS Personal Demographics Service (PDS). Due to an increase in digital prescription processing through the Electronic Prescription Service (EPS) during the COVID-19 pandemic, more patients were identified in 2020/21 and 2021/22 compared to previous years.

The NIC does not include other fees incurred by dispensing contractors, such as controlled drug fees or the single activity fee. The basic price is determined by the [Drug Tariff](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff) or by the manufacturer, wholesaler, or supplier of the product.

### 1.3. Time periods {.toc-ignore}

The UK had several lockdowns in response to the COVID-19 pandemic. The first lockdown was from March to July 2020, followed by a second from November to December 2020. The third lockdown began in January 2021, and some restrictions were lifted beginning in March 2021. The Omicron variant led to new measures starting in November 2021, but there was no lockdown. In February 2022, all domestic legal COVID-19 restrictions were lifted in England. A detailed [timeline of COVID-19 lockdowns](https://commonslibrary.parliament.uk/research-briefings/cbp-9068/) can be found on the UK parliament website. It’s important to note that these statistics can’t prove that lockdowns caused changes in prescribing figures. 

The data in this publication covers the period from April 2017 to December 2022. Q1, Q2, Q3, and Q4 refer to different periods within a financial year. You can find more information on the [Medicines Used in Mental Health - England webpage](https://www.nhsbsa.nhs.uk/statistical-collections/medicines-used-mental-health-england).

### 1.4. Patient identification {.toc-ignore}

The NHSBSA processes prescriptions but doesn’t always identify the NHS number of the patient. Table 1 shows the proportion of items for which a patient could be identified. This means that the data relating to patient counts only represents most, but not all, patients.

As the patient identification rates increased during the COVID-19 pandemic, it is worth noting that any increases in the number of identified patients is likely to be an overestimate of the actual increase in patient numbers. This is as the proportion of patients who could be identified has increased. Conversely, any decrease over the same period is likely to be an underestimate of the actual decrease.

These statistics do not include any information that can identify a specific person. You can find more information about how the NHSBSA protect personal information in the [confidentiality and access statement](https://www.nhsbsa.nhs.uk/policies-and-procedures). 

#### Table 1: The proportion of items for which an NHS number was recorded for listed BNF sections for previous 4 quarters

```{r pat_ident_rates, out.width="100%"}

patient_identification %>%

  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t",
                   columnDefs = list(list(orderable = FALSE,
                                          targets = "_all"),
                                     list(className = "dt-left", targets = 0:1),
                                     list(className = "dt-right", targets = 2:5)))
  )

```

Where the proportion of identified patients has increased, this could result in an overestimation of any increases in patients receiving prescribing for a BNF section. Where the proportion has fallen, this could result in an underestimate.
---

## 2. Results and commentary

### 2.1. Antidepressants

Antidepressant drugs are effective for treating moderate to severe depression. Antidepressant drugs are not generally advised for the routine treatment of mild depression, and psychological therapy should be considered initially. However, a trial of antidepressant therapy may be considered in cases resistant to psychological treatments or associated with other problems. The National Institute for Health and Care Excellence (NICE) has [released updated guidance](https://www.nice.org.uk/news/nice-draft-quality-standard-depression-adults-update-2023) to recommend adult patients stopping antidepressants should have their dose reduced in stages across time.

Antidepressant drugs are described in the BNF 68 section 4.3:

* BNF paragraph 4.3.1 - Tricyclic and related antidepressant drugs
* BNF paragraph 4.3.2 - Monoamine-oxidase inhibitors (MAOIs)
* BNF paragraph 4.3.3 - Selective serotonin re-uptake inhibitors (SSRIs)
* BNF paragraph 4.3.4 - Other antidepressant drugs

More information about these medicines can be found in section 4 of this summary.

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.3: Antidepressant drugs

```{r quarterly_003, out.width="100%"}

raw_data$quarterly %>%
  #added financial year filter now only using rolling years
  dplyr::filter(SECTION_CODE == "0403"& FINANCIAL_YEAR %!in% c("2015/2016", "2016/2017")) %>% 
  dplyr::group_by(FINANCIAL_QUARTER, SECTION_NAME, SECTION_CODE) %>% 
  dplyr::summarise(`Prescribed items` = sum(ITEM_COUNT),
            `Identified patients` = sum(PATIENT_COUNT),
            .groups = "drop") %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  dplyr::arrange(desc(measure)) %>% 
  group_chart_hc(
    x = FINANCIAL_QUARTER,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Financial quarter",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)
  
```
Source: [Quarterly Summary Tables - Quarterly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

In October to December 2022, prescribing of antidepressant drug items increased to 21.7 million items, 2.67% more than the same time last year. This continues the long-term trend for antidepressant prescribing, which has risen by 25% since 2017/18. 

An estimated 6.59 million identified patients were prescribed at least one antidepressant item between July to September 2022. This was not a notable change from the previous 3 months, but estimated identified patients are increasing along the same long-term trend as items.   

#### Figure 2: Number of prescribed items and identified patients by month - BNF 4.3: Antidepressant drugs

```{r monthly_0403, out.width="100%"}

chart_data$monthly %>% 
  #year month filter added for rolling years in narrative
  dplyr::filter(SECTION_CODE == "0403"& YEAR_MONTH>201703) %>% 
  dplyr::rename(`Prescribed items` = ITEM_COUNT,
         `Identified patients` = PATIENT_COUNT) %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  group_chart_hc(
    x = MONTH_START,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Month",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_xAxis(type = "datetime",
           dateTimeLabelFormats = list(month = "%b %y"),
           title = list(text = "Month")) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)



```
Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

Figure 2 shows antidepressant prescribing continues to have regular and seasonal patterns. Months with less dispensing days, such as February, tend to have less prescribing. Despite the month-to-month variation, antidepressants show a clear rise over time.

In the 12 months of 2022, there was a monthly average of 4.57 million identified patients receiving at least one antidepressant item. This was a small increase from the monthly average of 4.45 million identified patients in 2021.

There were 85.4 million antidepressant drug items prescribed in the 12 months January 2022 to December 2022, a 3.37% increase from 82.6 million items in 2021. 

### 2.2. Hypnotics and anxiolytics

Hypnotics are used to treat insomnia - a difficulty getting to sleep or staying asleep long enough to feel refreshed. Most hypnotics will sedate if given during the day. Anxiolytics are used in the treatment of anxiety states and if given at night will help to induce sleep. The drugs classified as hypnotics and anxiolytics are described in the BNF 68 section 4.1:

* BNF paragraph 4.1.1 - Hypnotics
* BNF paragraph 4.1.2 - Anxiolytics
* BNF paragraph 4.1.3 - Barbiturates

More information about these medicines can be found in section 4 of this summary.

#### Figure 3: Number of prescribed items and identified patients by financial quarter - BNF 4.1: Hypnotics and anxiolytics

```{r quarterly_0401, out.width="100%"}

raw_data$quarterly %>%
  dplyr::filter(SECTION_CODE == "0401"& FINANCIAL_YEAR %!in% c("2015/2016", "2016/2017")) %>% 
  dplyr::group_by(FINANCIAL_QUARTER, SECTION_NAME, SECTION_CODE) %>% 
  dplyr::summarise(`Prescribed items` = sum(ITEM_COUNT),
            `Identified patients` = sum(PATIENT_COUNT),
            .groups = "drop") %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  dplyr::arrange(desc(measure)) %>% 
  group_chart_hc(
    x = FINANCIAL_QUARTER,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Financial quarter",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)
  
```

Source: [Quarterly Summary Tables - Quarterly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

There were 3.44 million hypnotics and anxiolytics prescribed in Q3 of 2022/23. This was a 3.14% decrease from the same quarter a year ago, but similar to Q2 of 2022/23. Prescribing of hypnotics and anxiolytics has been decreasing since 2017/18, with 10.6% fewer items between Q3 of 2017/18 and Q3 of 2022/23.

Between October to December 2022, an estimated 1.03 million identified patients were prescribed at least one hypnotics and anxiolytics item. This was only a small decrease from the three months before. Identified patients have decreased by 12.7% between Q3 of 2017/18 and Q3 2022/23.  This is similar to the decrease in items over the same time.

#### Figure 4: Number of prescribed items and identified patients by month - BNF 4.1: Hypnotics and anxiolytics

```{r monthly_0401, out.width="100%"}

chart_data$monthly %>% 
  dplyr::filter(SECTION_CODE == "0401"& YEAR_MONTH>201703) %>% 
  dplyr::rename(`Prescribed items` = ITEM_COUNT,
         `Identified patients` = PATIENT_COUNT) %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  group_chart_hc(
    x = MONTH_START,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Month",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_xAxis(type = "datetime",
           dateTimeLabelFormats = list(month = "%b %y"),
           title = list(text = "Month")) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

In Figure 4, the month-to-month variation and regular patterns over time can be seen. Like the other BNF sections, there tends to be less prescribing for hypnotics and anxiolytics in months with fewer prescribing days. This is particularly visible in February.

Between January and December 2022, an average of 676,000 identified patients a month received at least one hypnotic and anxiolytics item. In 2021, there were an average of 691,000, so this was a small decrease. 

January to December 2022 also had a small decrease in hypnotics and anxiolytics items compared to 2021. There were 13.8 million items prescribed in 2022, and 14.1 million in 2021. 

### 2.3. Antipsychotics

These drugs are used to treat psychoses and related disorders. These drugs are classified and described under the BNF 68 section 4.2:

* BNF paragraph 4.2.1 - Antipsychotic drugs
* BNF paragraph 4.2.2 - Antipsychotic depot injections
* BNF paragraph 4.2.3 - Drugs for mania and hypomania

More information about these medicines can be found in section 4 of this summary.

#### Figure 5: Number of prescribed items and identified patients by financial quarter - BNF 4.2: Drugs used in psychoses and related disorders

```{r quarterly_0402, out.width="100%"}

raw_data$quarterly %>%
  dplyr::filter(SECTION_CODE == "0402"& FINANCIAL_YEAR %!in% c("2015/2016", "2016/2017")) %>%
  dplyr::group_by(FINANCIAL_QUARTER, SECTION_NAME, SECTION_CODE) %>% 
  dplyr::summarise(`Prescribed items` = sum(ITEM_COUNT),
            `Identified patients` = sum(PATIENT_COUNT),
            .groups = "drop") %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  dplyr::arrange(desc(measure)) %>% 
  group_chart_hc(
    x = FINANCIAL_QUARTER,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Financial quarter",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)
  
```

Source: [Quarterly Summary Tables - Quarterly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

Antipsychotics items increased to 3.38 million items prescribed in October to December 2022. This was only a very small increase from the previous three months and the same three months in 2021. Antipsychotics items show a long-term trend of increasing over time, with 12.1% more items in Q3 2022/23 than in Q3 2017/18. 

October to December 2022 had an estimated 646,000  identified patients prescribed at least one antipsychotic item. This was not a notable change from the 640,000 identified patients in the previous three months or 637,000 identified patients in October to December 2021. Between Q3 2017/18 and Q3 2022/23, identified patients increased by 11.0%, similar to the increase in items. 

#### Figure 6: Number of prescribed items and identified patients by month - BNF 4.2: Drugs used in psychoses and related disorders

```{r monthly_0402, out.width="100%"}

chart_data$monthly %>% 
  dplyr::filter(SECTION_CODE == "0402"& YEAR_MONTH>201703) %>% 
  dplyr::rename(`Prescribed items` = ITEM_COUNT,
         `Identified patients` = PATIENT_COUNT) %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  group_chart_hc(
    x = MONTH_START,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Month",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_xAxis(type = "datetime",
           dateTimeLabelFormats = list(month = "%b %y"),
           title = list(text = "Month")) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

The regular and seasonal patterns over time for antipsychotics can be seen in Figure 6. There tends to be less prescribing for hypnotics and anxiolytics in months with fewer prescribing days. This is particularly visible in February, the same as with other BNF sections. The month-to-month changes for antispychotics items appear more volatile than the changes in identified patient numbers.

For January to December 2022, the monthly average of identified patients who received at least one antipsychotic item was 493,000. Unlike the other BNF sections, this was almost unchanged from 492,000 monthly average identified patients in 2021.

From January to December 2022, there were 13.3 million antipsychotic items prescribed, only a small increase from 2021 which had 13.2 million.

### 2.4. Central nervous system (CNS) stimulants and drugs used for ADHD 

CNS stimulants and drugs used for ADHD are described in the BNF 68 section 4.4: 

* BNF chemical substance 0404000D0 - Caffeine
* BNF chemical substance 0404000E0 - Caffeine citrate
* BNF chemical substance 0404000L0 - Dexamfetamine sulfate
* BNF chemical substance 0404000M0 - Methylphenidate hydrochloride
* BNF chemical substance 0404000R0 - Modafinil
* BNF chemical substance 0404000S0 - Atomoxetine hydrochloride
* BNF chemical substance 0404000T0 - Dexmethylphenidate hydrochloride
* BNF chemical substance 0404000U0 - Lisdexamfetamine dimesylate
* BNF chemical substance 0404000V0 - Guanfacine
* BNF chemical substance 0404000W0 - Pitolisant hydrochloride

More information about these medicines can be found in section 4 of this summary.

#### Figure 7: Number of prescribed items and identified patients by financial quarter - BNF 4.4: CNS stimulants and drugs used for ADHD

```{r quarterly_0404, out.width="100%"}

raw_data$quarterly %>%
  dplyr::filter(SECTION_CODE == "0404"& FINANCIAL_YEAR %!in% c("2015/2016", "2016/2017")) %>%
  dplyr::group_by(FINANCIAL_QUARTER, SECTION_NAME, SECTION_CODE) %>% 
  dplyr::summarise(`Prescribed items` = sum(ITEM_COUNT),
            `Identified patients` = sum(PATIENT_COUNT),
            .groups = "drop") %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  dplyr::arrange(desc(measure)) %>% 
  group_chart_hc(
    x = FINANCIAL_QUARTER,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Financial quarter",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)
  
```
Source: [Quarterly Summary Tables - Quarterly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

Prescribing of CNS stimulants and drugs for ADHD continues to show an upwards trend, with 646,000 items prescribed in October to December 2022. This was an increase on 607,000 items in the previous three months, and from 548,000 items in October to December 2021. Since October to December 2017 there has been a 69.1% increase, with an estimated 264,000 more items prescribed in October to December 2022.

Both items and identified patients are increasing over time. An estimated 180,000 identified patients were prescribed at least one CNS stimulants and drugs used for ADHD item in October to December 2022. This was a noticeable increase from 170,000 in the previous three months, and from 149,000 in October to December 2021. There has been an overall rise of 82,600 identified patients since October to December 2017, an 84.8% increase. 

CNS stimulants and drugs used for ADHD have seen a larger rise in the proportion of patients identified than other BNF sections. This may result in an overestimate of the actual increase. More information about patient identification can be found in the introduction.

#### Figure 8: Number of prescribed items and identified patients by month - BNF 4.4: CNS stimulants and drugs used for ADHD

```{r monthly_0404, out.width="100%"}

chart_data$monthly %>% 
  dplyr::filter(SECTION_CODE == "0404"& YEAR_MONTH>201703) %>% 
  dplyr::rename(`Prescribed items` = ITEM_COUNT,
         `Identified patients` = PATIENT_COUNT) %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  group_chart_hc(
    x = MONTH_START,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Month",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_xAxis(type = "datetime",
           dateTimeLabelFormats = list(month = "%b %y"),
           title = list(text = "Month")) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

Figure 8 also fits the pattern from the other BNF sections of showing regular and seasonal patterns. There tends to be less prescribing for hypnotics and anxiolytics in months with fewer prescribing days. This is particularly visible in February. CNS stimulants and drugs used in ADHD also have a visible decline every August in recent years.

The monthly average of identified patients receiving at least one CNS stimulants and drugs used for ADHD item over January to December 2021 was 116,000. In 2021, there was a monthly average of 98,900, so this is an increase. 

There was also an increase in items prescribed, from 2.04 million items in January to December 2021, to 2.42 million in 2022.

### 2.5. Drugs for dementia

Dementia is a progressive clinical syndrome characterised by a range of cognitive and behavioural symptoms that can include memory loss, problems with reasoning and communication, a change in personality, and a reduced ability to carry out daily activities such as washing or dressing. Alzheimer’s disease is the most common type of dementia. Other common types of dementia include vascular dementia (where dementia is due to cerebrovascular disease), dementia with Lewy bodies (where dementia is due to protein deposits in the brain), mixed dementia, and frontotemporal dementia (where dementia is due to progressive nerve loss in either the frontal or temporal lobes of the brain).

Drugs for dementia are described in the BNF 68 section 4.11:

* BNF chemical substance 0411000D0 - Donepezil hydrochloride
* BNF chemical substance 0411000E0 - Rivastigmine
* BNF chemical substance 0411000F0 - Galantamine
* BNF chemical substance 0411000G0 - Memantine hydrochloride

More information about these medicines can be found in section 4 of this summary.

#### Figure 9: Number of prescribed items and identified patients by financial quarter - BNF 4.11: Drugs for dementia

```{r quarterly_0411, out.width="100%"}

raw_data$quarterly %>%
  dplyr::filter(SECTION_CODE == "0411"& FINANCIAL_YEAR %!in% c("2015/2016", "2016/2017")) %>%
  dplyr::group_by(FINANCIAL_QUARTER, SECTION_NAME, SECTION_CODE) %>% 
  dplyr::summarise(`Prescribed items` = sum(ITEM_COUNT),
            `Identified patients` = sum(PATIENT_COUNT),
            .groups = "drop") %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  dplyr::arrange(desc(measure)) %>% 
  group_chart_hc(
    x = FINANCIAL_QUARTER,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Financial quarter",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)
  
```

Source: [Quarterly Summary Tables - Quarterly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

There were 1.08 million drugs for dementia prescribed in October to December 2022. This is a small rise from 1.06 million items in the previous three months. Although items have increased overall by 9.47% since October to December 2017, drugs for dementia have not shown a consistent long term trend over this time. 

An estimated 242,000 identified patients in October to December 2022 were prescribed at least one drugs for dementia item. This was only a small increase on identified patients in the previous three months. Like items, identified patients have not shown a consistent trend.

#### Figure 10: Number of prescribed items and identified patients by month - BNF 4.11: Drugs for dementia

```{r monthly_0411, out.width="100%"}

chart_data$monthly %>% 
  dplyr::filter(SECTION_CODE == "0411"& YEAR_MONTH>201703) %>% 
  dplyr::rename(`Prescribed items` = ITEM_COUNT,
         `Identified patients` = PATIENT_COUNT) %>% 
  tidyr::pivot_longer(cols = c(`Identified patients`,`Prescribed items`),
               names_to = "measure",
               values_to = "value") %>% 
  dplyr::mutate(value = signif(value, 3)) %>% 
  group_chart_hc(
    x = MONTH_START,
    y = value,
    group = measure,
    type = "line",
    marker = FALSE,
    dlOn = FALSE,
    xLab = "Month",
    yLab = "Volume",
    title = ""
  ) %>% 
  hc_xAxis(type = "datetime",
           dateTimeLabelFormats = list(month = "%b %y"),
           title = list(text = "Month")) %>% 
  hc_tooltip(enabled = TRUE,
             shared = TRUE,
             sort = TRUE) %>% 
  hc_legend(enabled = TRUE)

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_dec22_v001.xlsx)

The monthly average number of identified patients from January to December 2021 for drugs for dementia items was 194,000, a small rise from 188,000 in 2021.

Over January to December 2022, 4.19 million drugs for dementia items were prescribed. This was also a small increase from 2021, when there were 4.03 million items.

## 3. Prescribing during the COVID-19 pandemic 

The number of expected prescription items for the below sections during this period have been calculated by extrapolating figures for March 2020 to `r max_filter_12_months` according to:

* the trend in prescription items for each section between April 2015 and February 2020
* the number of dispensing days that occur in each month
* the typical differences between each month of the year observed between April 2015 and February 2020


### 3.1. Antidepressant Prescribing during the COVID-19 pandemic {.toc-ignore}

#### Figure 11: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.3: Antidepressant drugs

```{r covid_0403_items, out.width="100%"}

 model_data_old %>%
  covid_model() %>% 
  filter(SECTION_CODE == "0403", YEAR_MONTH > 202002) %>%
  covid_chart_hc(title = "")

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_sep22_v001.xlsx)

`r covpar_0403` However, this increase in prescribed items does not fall outside of the expected range of values for the period.

### 3.2. Hypnotics and anxiolytics prescribing during the COVID-19 pandemic {.toc-ignore}

#### Figure 12: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.1: Hypnotics and anxiolytics

```{r covid_0401_items, out.width="100%"}

model_data_old %>%
  covid_model() %>% 
  filter(SECTION_CODE == "0401", YEAR_MONTH > 202002) %>%
  covid_chart_hc(title = "")

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_sep22_v001.xlsx)

`r covpar_0401` However, this increase in prescribed items does not fall outside of the expected range of values for the period.

### 3.3. Drugs used in psychoses and related disorders prescribing during the COVID-19 pandemic {.toc-ignore}

#### Figure 13: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.2: Drugs used in psychoses and related disorders

```{r covid_0402_items, out.width="100%"}

model_data_old %>%
  covid_model() %>% 
  filter(SECTION_CODE == "0402", YEAR_MONTH > 202002) %>%
  covid_chart_hc(title = "")

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_sep22_v001.xlsx)

`r covpar_0402` This small decrease in prescribed items is in line with the expected range of values for the period.

### 3.4. CNS stimulants and drugs used for ADHD prescribing during the COVID-19 pandemic {.toc-ignore}

#### Figure 14: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.4: CNS stimulants and drugs used for ADHD

```{r covid_0404_items, out.width="100%"}

model_data_old %>%
  covid_model() %>% 
  filter(SECTION_CODE == "0404", YEAR_MONTH > 202002) %>%
  covid_chart_hc(title = "")

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_sep22_v001.xlsx)

`r covpar_0404` This increase in prescribed items falls outside of the expected range of values for the overall period. Since November 2021 and including all 3 months of quarter 2 2022/23, there have been 11 consecutive months above the number of items expected based on historical trends. 

### 3.5. Drugs for dementia prescribing during the COVID-19 pandemic {.toc-ignore}

#### Figure 15: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.11: Drugs for dementia

```{r covid_0411_items, out.width="100%"}

model_data_old %>%
  covid_model() %>% 
  filter(SECTION_CODE == "0411", YEAR_MONTH > 202002) %>%
  covid_chart_hc(title = "")

```

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_quarterly_sep22_v001.xlsx)

`r covpar_0411` This decrease also falls outside of the expected range of values, showing a statistically significant decrease in prescribing of drugs for dementia in this period when compared to expected values.

---

## 4. Background

### 4.1. Antidepressant drugs {.toc-ignore}

Antidepressant drugs are licensed to treat major depression. Health professionals use the words depression, depressive illness or clinical depression to refer to depression. It is a serious illness and very different from the common experience of feeling unhappy or fed up for a short period of time. Depressed people may have feelings of extreme sadness that can last for a long time. These feelings are severe enough to interfere with daily life, and can last for weeks, months or years, rather than days.

It should be noted that antidepressant drugs are used for indications other than depression, for example migraine, chronic pain, myalgic encephalomyelitis (ME), or a range of other conditions. Clinical indication isn't captured by the NHSBSA. Therefore, the statistics on these drugs do not relate solely to prescribing for depression.

You can find more information about [depression](https://www.nhs.uk/conditions/clinical-depression/) on the NHS website.

### 4.2. Hypnotics and anxiolytics {.toc-ignore}

Hypnotics and anxiolytics are used to treat insomnia and anxiety respectively.

Insomnia is difficulty getting to sleep or staying asleep for long enough to feel refreshed in the morning, despite there being enough opportunity to sleep. The most common problem with insomnia is difficult falling asleep (sleep-onset insomnia). An insomniac may also experience:

* waking in the night
* not feeling refreshed after sleep and not being able to function normally during the day
* feeling irritable and tired and finding it difficult to concentrate
* waking when they have been disturbed from sleep by pain or noise
* waking early in the morning

Anxiety is a feeling of unease, such as worry or fear, which can be mild or severe. Everyone experiences feelings of anxiety at some point in their life and feeling anxious is sometimes perfectly normal. However, people with generalised anxiety disorder (GAD) find it hard to control their worries. Their feelings of anxiety are more constant and often affect their daily life. There are several conditions for which anxiety is the main symptom. Panic disorder, phobias and post-traumatic stress disorder can all cause severe anxiety.

You can find more information about [insomnia](https://www.nhs.uk/conditions/insomnia/) and [anxiety](https://www.nhs.uk/conditions/generalised-anxiety-disorder/) from the NHS website.

### 4.3. Drugs used in psychoses and related disorders {.toc-ignore}

People experiencing psychoses may report hallucinations (seeing or hearing things that other people cannot see or hear) or delusional beliefs (believing things that are not actually true), and may exhibit personality changes and thought disorder. As well as this, they may experience difficulties with social interaction and impairment in carrying out daily life activities.

You can find more information about [psychoses and related disorders](https://www.nhs.uk/conditions/psychosis/) on the NHS website.

### 4.4. Central nervous system (CNS) stimulants and drugs used for ADHD  {.toc-ignore}

Attention deficit hyperactivity disorder (ADHD) and attention deficit disorder (ADD) refer to a range of problem behaviours associated with poor attention span. These may include impulsiveness and hyperactivity, as well as inattentiveness; behaviours that often prevent children and adults from learning and socialising. ADHD is sometimes referred to as hyperkinetic disorder (HD).

You can find more information about [ADHD](https://www.nhs.uk/conditions/attention-deficit-hyperactivity-disorder-adhd/) on the NHS website.

### 4.5. Drugs for dementia {.toc-ignore}

Dementia is a disease that leads to progressive loss of brain function typified by memory loss, confusion, speech difficulties and problems in understanding. There are over 100 different types of dementia. The most common forms are:

* Alzheimer's disease
* Vascular dementia
* Dementia with Lewy bodies
* Pick's disease
* Huntington's disease
* Alcohol-related dementia
* HIV/AIDS related dementia

Dementia mainly affects older people but can also occur in people as young as thirty due to either alcohol abuse or HIV/AIDS. No cure for dementia currently exists. However, drugs may slow the rate of decline or in some patients make a small improvement in symptoms. Despite this, disease progression is inevitable.

You can find more information about the [various types of dementia](https://www.nhs.uk/conditions/dementia/) on the NHS website.

### 4.6. Related statistics and resources

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS Digital mental health data hub] (https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide useful context. These include reports on children and young people's mental health surveys, a mental health bulletin, dementia assessment and referral statistics, and more.

---

## 5. About these statistics

This publication is part of a series by the NHSBSA. This quarterly summary is intended to supplement the [annual summary statistics](https://www.nhsbsa.nhs.uk/statistical-collections/medicines-used-mental-health-england/medicines-used-mental-health-england-201516-202122) released in each year and previous quarterly summaries. To learn more about how this series compares to our other publications, you can view the Official Statistics guidance table on our [statistical collections page](https://www.nhsbsa.nhs.uk/statistical-collections). 

### 5.1. Patient counts {.toc-ignore}

The patient counts shown in these statistics should only be analysed at the level at which they are presented. Adding together any patient counts is likely to result in an overestimate of the number of patients. A patient will be included, or counted, in each category or time period in which they received relevant prescriptions. For example, if a patient received a prescription item for an antidepressant drug in Q1 2020/21 and another in Q2 2020/21, then adding together those totals would count that patient twice. For the same reason, data on patient counts for different BNF sections should not be added together.

### 5.2. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity, in most cases to three significant figures. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

### 5.3. Trends during COVID-19 {.toc-ignore}

The number of items for each BNF section in the 31-month period March 2020 to `r max_filter_12_months` were compared to the number of items expected to be prescribed based on trends from the 'pre-COVID-19' period. The monthly data for April 2015 to February 2020 was used to extrapolate expected figures for March 2020 to `r max_filter_12_months`. These figures were extrapolated according to:

* the trends between April 2015 and February 2020
* the number of dispensing days that occur in each month
* the typical differences between each month of the year observed between April 2015 and February 2020

This model states expected figures if trends had continued as before but doesn't confirm causation of any differences. Factors other than COVID-19 may have influenced prescribing, such as:

*	changes in patient behaviour, demographics, or morbidity
*	external environmental factors that could drive prescribing
*	any (and all) policy initiatives and guidance
*	other local priorities and resources

### 5.4. Averages {.toc-ignore}

Where this document refers to 'average', this is the mean unless otherwise stated. This is calculated by adding the number of items/patients/months together and dividing this by the number of items/patients/months.

### 5.5. Planned changes to this publication {.toc-ignore}

This is an experimental official statistic release. Experimental statistics are newly developed or innovative statistics. These are published so that users and stakeholders can be involved in the assessment of their suitability and quality at an early stage. We will regularly be reviewing the methodology used within the statistics.

---

## 6. Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: nhsbsa.accessibility@nhs.net

**Phone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Contact us’ section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---

## 7. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics. 

You can complete a [short survey about this publication](https://online1.snapsurveys.com/interview/caf0d685-0500-48d2-8b96-1c434a956819) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** nhsbsa.statistics@nhs.net

**You can also write to us at:**

NHSBSA - Statistics  
NHS Business Services Authority  
Stella House  
Goldcrest Way  
Newburn Riverside  
Newcastle upon Tyne  
NE15 8NY

**Responsible statistician:** Grace Libby

</main>
